---
title: "Akula-DATA607-Week05-Homework"
author: "Pavan Akula"
date: "March 01, 2017"
output: html_document
---

This assignment is about using R to tidy the given data, moving between "wide" and "long" formats to best support our analysis. Following libraries will be used "base R", tidyr and dplyr to complete the assignment.

Load the libraries
```{r , echo=T, message=F}
library(tidyr)    # gather() and spread() functions are used to tidy the data
library(dplyr)    # functions select(), summarise() and filter() are from this library
library(stringr)  # used for pattern matching and data cleansing 
```

Sample CSV file is created using data instructions provided in **_Assignment Tidying and Transforming Data.pdf._**

```{r , echo=T}
localDir <- getwd()
flightsFile <- paste(str_trim(localDir),"//Flights.csv",sep="")

# Data into a data frame
flights.raw <- read.csv(flightsFile, header = T)


# Rename the column names accordingly
flights.raw <- rename(flights.raw, `Los Angeles` = `Los.Angeles`, `San Diego` =`San.Diego`, `San Francisco` = `San.Francisco`, Airline = X, ArrivalInfo = `X.1`)

flights.raw

```

```{r , echo=T}
# Remove rows that are empty.
# flights.raw <- flights.raw %>% filter(`Los Angeles` != "NA")

flights.raw <- flights.raw %>%
  filter(!is.na(`Los Angeles`))

flights.raw

```

```{r , echo=T}
# Format data
# 2 rows make up 1 row, following command populate Airline name in second row.
# Fill in missing data.

flights.raw[seq(2,nrow(flights.raw),2),1] <- flights.raw[seq(1,nrow(flights.raw),2),1]

flights.raw

```

```{r , echo=T}
# tidy data
# flights.raw dataset has 3 variables(Airline, ArrivalInfo, Cities) and 2 observations(on time, delayed).
# 1. Airline has 2 possible values (Alaska, AM West)
# 2. ArrivalInfo has 2 possible values(on time and delayed)
# 3. Cities has 5 possible values(Los Angeles, Phoenix, San Diego, San Francisco, Seattle)
# 4. Observations have 20 values (497,221,212,503,1841,62,12,20,102,305,694,4840,383,320,201,117,415,65,129,61)

flights.table1 <- flights.raw %>% 
  gather(City, Count, -Airline, -ArrivalInfo, na.rm = TRUE)

# A tidy version of the flights.raw data looks like this
flights.table1
```

Data set is mostly tidy except for **_ArrivalInfo_** column. **_ArrivalInfo_** stores the names of variables.

```{r , echo=T}
# Expand rows into columns

flights.table2 <- flights.table1 %>% 
  spread(ArrivalInfo, Count) %>%
  select (Airline, City,`on time`, delayed)

flights.table2
```

Analysis of the data.

```{r , echo=T}
# Add a new column to find total flights by airline using mutate function

flights.table3 <- flights.table2 %>%
  mutate(TotalFlights = flights.table2$`on time`+flights.table2$delayed , 
         PercentageOfOnTime = round(flights.table2$`on time` * 100/TotalFlights,2) ,
         PercentageOfDelay = round(flights.table2$delayed * 100/TotalFlights,2)
        ) %>%
  filter(City %in% c("Los Angeles","Phoenix")) %>%
  select(Airline,City,PercentageOfOnTime,PercentageOfDelay) %>%
  arrange(City,Airline,PercentageOfDelay)


# By looking at the data for cities Los Angeles and Phoenix both airlines perform close enough. However, ALASKA airlines is better with on time arrival.
flights.table3
```

```{r , echo=T}
# Data for San Diego, San Francisco and Seattle

flights.table3 <- flights.table2 %>%
  mutate(TotalFlights = flights.table2$`on time`+flights.table2$delayed , 
         PercentageOfOnTime = round(flights.table2$`on time` * 100/TotalFlights,2) ,
         PercentageOfDelay = round(flights.table2$delayed * 100/TotalFlights,2)
        ) %>%
  filter(City %in% c("San Diego","San Francisco","Seattle")) %>%
  select(Airline,City,PercentageOfOnTime,PercentageOfDelay) %>%
  arrange(City,Airline,PercentageOfDelay)


# By looking at the data for cities San Diego, San Francisco and Seattle, ALASKA airlines is better with on time arrival and also has less delays.
flights.table3
```

Additional analysis, Seattle airport is used as example.

```{r, echo=T, warning=F}
# Data for Seattle

flights.table3 <- flights.table1 %>% 
  filter(City == "Seattle") %>%
  select(Airline,ArrivalInfo,Count) %>%
  spread(ArrivalInfo,Count)

# Bind row totals

flights.table3$Totals <- rowSums(flights.table3[,-1] )

# Bind column totals

flights.table3 %<>%
  summarise(Airline = "Totals",
            delayed = sum(delayed),
            `on time` = sum(`on time`),
            Totals = sum(Totals)) %>%
  bind_rows(flights.table3,.)

flights.table3
```
Assuming ALASKA and AM WEST are the olny airlines flying to Seattle,

1. Given random flight being delayed at Seattle airport, probability that it belongs to ALASKA airlines : **`r round(flights.table3[1,2]/flights.table3[1,4],2)`**

2. Given random flight being delayed at Seattle airport, probability that it belongs to AM WEST airlines : **`r round(flights.table3[2,2]/flights.table3[2,4],2)`**

3. Given random flight being on time at Seattle airport, probability that it belongs to ALASKA airlines : **`r round(flights.table3[1,3]/flights.table3[1,4],2)`**

4. Given random flight being on time at Seattle airport, probability that it belongs to AM WEST airlines : **`r round(flights.table3[2,3]/flights.table3[2,4],2)`**

5. Of all the airlines flying to Seattle, percentage of ALASKA airlines flights : **`r round(flights.table3[1,4] * 100/flights.table3[3,4],2)`**

6. Of all the airlines flying to Seattle, percentage of AM WEST airlines flights : **`r round(flights.table3[2,4] * 100 /flights.table3[3,4],2)`**

###Conclusion:
**_It is fair to say ALASKA airlines has better on time flights over AM WEST airlines._**


# Last week's project without using sqldf library


Libraries used for this project.

```{r , echo=TRUE}
library(readr)      # used for reading the text file
library(stringr)    # used for extracting text using regular expressions
library(tidyr)      # replaced sqldf 
library(dplyr)      # replaced sqldf
library(utils)      # used for generating CSV file
```

Load *_Chess Tournament Cross Table_* file. 

```{r , echo=TRUE}
# Get local directory and load the file

localDir <- getwd()
chessFile <- paste(str_trim(localDir),"//tournamentinfo.txt",sep="")

# Skip top 3 lines as it contains header info

chess.text<-read_lines(chessFile,skip = 3)
head(chess.text,10)
```

There are total of 64 players. File needs formatting in order to perform calculations and generate CSV output.

```{r , echo=TRUE}
# First level of formatting removes hypen("-") characters

chess.text.level1 <- gsub("-","",chess.text)
head(chess.text.level1, 6)
```
```{r , echo=TRUE}
# Second level of formatting removes empty lines
# This makes a continous data without newline characters 

chess.text.level2 <- paste(unlist(chess.text.level1), collapse = " ")
head(chess.text.level2,10)
```

```{r , echo=TRUE}
# Add extra characters at the end of text data. This helps read complete data. 
# If extra characters are not added, last line will be skipped. 

chess.text.level2 <- paste(chess.text.level2,"_*_*_*_")

# Extract the text using pattern repeating 180 characters
# Idea is to combine 2 lines of each player's info into single line
# Once combined, information about single player takes 180 characters

pattern <- "[[:print:]]{180}"
chess.text.level3 <- unlist(str_extract_all(chess.text.level2, pattern = pattern))
head(chess.text.level3,3)
```

```{r , echo=TRUE}
# Each player has about 20 columns of information. Split the single string into 20 columns

chess.data <- data.frame(unlist(str_split_fixed(chess.text.level3, "\\|", 21)))
head(chess.data, 5)
```

```{r , echo=TRUE}
# Using regular expressions extract the pre match rating for each player

pattern = "(R:\\s+[[:digit:]]{3,4})"
chess.preGameRating <- unlist(str_extract(string = chess.data[,12], pattern = pattern))

# Get the pre-game rating
pattern = "[[:digit:]]{3,4}"
chess.preGameRating <- str_extract(string = chess.preGameRating, pattern = pattern)
chess.preGameRating <- data.frame(chess.data, preGameRating = chess.preGameRating)

head(chess.preGameRating, 5)
```

```{r , echo=TRUE}
# Format the column type

chess.preGameRating$X1 <- as.numeric(as.character(chess.preGameRating$X1))
chess.preGameRating$preGameRating <- as.numeric(as.character(chess.preGameRating$preGameRating))

```

```{r , echo=TRUE}
# Extract the information about opponent for each round
# Opponent player info is left as "NA" values if player did not play or got bye

# Get opponet playerID
pattern = "[[:digit:]]{1,2}"

chess.preGameRating$G1opponent = as.numeric(str_match(string = chess.preGameRating$X4, pattern = pattern))
chess.preGameRating$G2opponent = as.numeric(str_match(string = chess.preGameRating$X5, pattern = pattern))
chess.preGameRating$G3opponent = as.numeric(str_match(string = chess.preGameRating$X6, pattern = pattern))
chess.preGameRating$G4opponent = as.numeric(str_match(string = chess.preGameRating$X7, pattern = pattern))
chess.preGameRating$G5opponent = as.numeric(str_match(string = chess.preGameRating$X8, pattern = pattern))
chess.preGameRating$G6opponent = as.numeric(str_match(string = chess.preGameRating$X9, pattern = pattern))
chess.preGameRating$G7opponent = as.numeric(str_match(string = chess.preGameRating$X10, pattern = pattern))

# Get required data
chess.data <- chess.preGameRating %>% 
  select(playerID = as.numeric(X1),playerName = X2, playerPoints = as.numeric(X3), playerState = X11, preGameRating, G1opponent,G2opponent,G3opponent,G4opponent,G5opponent,G6opponent,G7opponent)

# Convert data from factors to character

chess.data$playerName <- as.character(chess.data$playerName)
chess.data$playerState <- as.character(chess.data$playerState)
chess.data$playerPoints <- as.character(chess.data$playerPoints)

# Remove extra spaces

trim <- function (x) gsub("^\\s+|\\s+$", "", x)
chess.data$playerState <- trim(chess.data$playerState)
chess.data$playerName <- trim(chess.data$playerName)
chess.data$playerPoints <- trim(chess.data$playerPoints)

# Player 62 did not play 2, 3 or 4 games
tail(chess.data,10)
```

```{r , echo=TRUE}
# Tidy all data, convert all the opponet values into rows
# opponent is new column with values(G1opponent, G2opponent, ... G7opponent)
# n is the column that stored values for opponetID

chess.tidy.data <- chess.data %>% 
  gather(opponent, n, -playerID, -playerName, -playerState, -preGameRating, -playerPoints, na.rm = TRUE) %>%
  arrange(playerID)

# Rename the column
chess.tidy.data <-  rename(chess.tidy.data, opponentID = n)
head(chess.tidy.data,15)

# Remove the column that is not being used.
chess.tidy.data$opponent <- NULL
```

```{r , echo=TRUE}
# Following code joins chess.tidy.data and chess.data using "opponentID" = "playerID" and extracts opponentID pre-game rating.
# It summarises data using playerID and playerName, this gives sum of opponent ratings and number games played.
# Finally uses mutate to calculate average opponent ratings
# And saves required data to chess.rating data frame

chess.rating <- chess.tidy.data %>% inner_join(chess.data, by = c("opponentID" = "playerID")) %>%
  select (playerID, playerName = playerName.x, opponentID, opponentPlayer = playerName.y , opponentRating = as.numeric(preGameRating.y)) %>%
  group_by (playerID,playerName) %>% 
  summarise(opponentRatingSum = sum(opponentRating), numGames = n()) %>% 
  inner_join(chess.data, by = ("playerID")) %>% 
  select (playerID, playerName = playerName.x, playerPoints, playerState, preGameRating, opponentRatingSum, numGames) %>% 
  mutate(avgOppPreGameRating = round(opponentRatingSum / numGames,0)) %>% 
  select (playerID, playerName, playerPoints, playerState, preGameRating, avgOppPreGameRating)


head(chess.rating,10)
```

```{r, echo=TRUE}
# Generate CSV files

chessOutFile <- paste(str_trim(localDir),"//chessOutFile.csv",sep="")
write.table(chess.rating, file = chessOutFile, row.names = FALSE, sep = ",")
```